\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2015-05-21 doc/USER_MANUAL/02_getting_started.tex
updates manual for better wiki display

To download the SPECFEM3D\_Cartesian software package, type this:
\begin{verbatim}
git clone --recursive --branch devel https://github.com/geodynamics/specfem3d.git
\end{verbatim}
%
We recommend that you add {\texttt{ulimit -S -s unlimited}} to your
{\texttt{.bash\_profile}} file and/or {\texttt{limit stacksize
unlimited }} to your {\texttt{.cshrc}} file to suppress any potential
limit to the size of the Unix stack.\\

Then, to configure the software for your system, run the \texttt{configure}
shell script. This script will attempt to guess the appropriate configuration
values for your system. However, at a minimum, it is recommended that
you explicitly specify the appropriate command names for your Fortran
compiler and MPI package (another option is to define FC, CC and MPIF90
in your .bash\_profile or your .cshrc file):
\begin{verbatim}
./configure FC=ifort MPIFC=mpif90
\end{verbatim}
Note that MPI must be installed with MPI-IO enabled because parts of SPECFEM3D perform I/Os through MPI-IO.

Before running the \texttt{configure} script, you should probably
edit file \texttt{flags.guess} to make sure that it contains the best
compiler options for your system. Known issues or things to check
are:
\begin{description}
\item [{\texttt{Intel ifort compiler}}] See if you need to add \texttt{-assume byterecl} for your machine. \textbf{In the case of that compiler, we have noticed that versions dot zero sometimes have bugs or issues that can lead to wrong results when running the code, thus we \emph{strongly} recommend using versions dot one or above (for instance version 13.1 instead of 13.0, version 14.1 instead of 14.0 and so on)}.
\item [{\texttt{IBM compiler}}] See if you need to add \texttt{-qsave}
or \texttt{-qnosave} for your machine.
\item [{\texttt{Mac OS}}] You will probably need to install \texttt{XCODE}.
\end{description}
When compiling on an IBM machine with the \texttt{xlf} and \texttt{xlc}
compilers, we suggest running the \texttt{configure} script with the
following options:\\

\noindent \texttt{./configure FC=xlf90\_r MPIFC=mpif90 CC=xlc\_r CFLAGS=\textquotedbl{}-O3
-q64\textquotedbl{} FCFLAGS=\textquotedbl{}-O3 -q64\textquotedbl{}
-with-scotch-dir=...}.\\

@@

The SCOTCH library \citep{PeRo96} provides efficent static mapping,
graph and mesh partitioning routines. SCOTCH is a free software package
developed by Fran\c{c}ois Pellegrini et al. from LaBRI and INRIA in Bordeaux,
France, downloadable from the web page \url{https://gforge.inria.fr/projects/scotch/}.
In case no SCOTCH libraries can be found on the system, the configuration
will bundle the version provided with the source code for compilation.
The path to an existing SCOTCH installation can to be set explicitly
with the option \texttt{-{}-with-scotch-dir}. Just as an example:
\begin{verbatim}
./configure FC=ifort MPIFC=mpif90 --with-scotch-dir=/opt/scotch
\end{verbatim}
If you use the Intel ifort compiler to compile the code, we recommend
that you use the Intel icc C compiler to compile Scotch, i.e., use:
\begin{verbatim}
./configure CC=icc FC=ifort MPIFC=mpif90
\end{verbatim}

@@

The DO-loop using OpenMP threads has a SCHEDULE property. The OMP\_SCHEDULE
environment variable can set the scheduling policy of that DO-loop.
Tests performed by Marcin Zielinski at SARA (The Netherlands) showed
that often the best scheduling policy is DYNAMIC with the size of
the chunk equal to the number of OpenMP threads, but most preferably
being twice as the number of OpenMP threads (thus chunk size = 8 for
4 OpenMP threads etc). If OMP\_SCHEDULE is not set or is empty, the
DO-loop will assume generic scheduling policy, which will slow down
the job quite a bit.

@@

\noindent Edit file \texttt{flags.guess} and put this for \texttt{FLAGS\_CHECK}:\\

\noindent \texttt{-g -qfullpath -O2 -qsave -qstrict -qtune=qp -qarch=qp
-qcache=auto -qhalt=w}\\
 \texttt{-qfree=f90 -qsuffix=f=f90 -qlanglvl=95pure -Q -Q+rank,swap\_all
-Wl,-relax}\\

@@

\noindent Then, to configure the code, type this:\\

\noindent \texttt{./configure FC=bgxlf90\_r MPIFC=mpixlf90\_r CC=bgxlc\_r
LOCAL\_PATH\_IS\_ALSO\_GLOBAL=true}\\

@@

\noindent To compile the code on an IBM BlueGene, Laurent L\'eger from
IDRIS, France, suggests the following: compile the code with

\noindent \texttt{ FLAGS\_CHECK=\textquotedbl{}-O3 -qsave -qstrict
-qtune=auto -qarch=450d -qcache=auto }~\\
 \texttt{ -qfree=f90 -qsuffix=f=f90 -g -qlanglvl=95pure -qhalt=w -Q
-Q+rank,swap\_all -Wl,-relax\textquotedbl{}}

@@

\noindent Also, \texttt{AR=ar, ARFLAGS=cru} and \texttt{RANLIB=ranlib}
are hardwired in all \texttt{Makefile.in} files by default, but to
cross-compile on BlueGene/P one needs to change these values to \texttt{AR=bgar,
ARFLAGS=cru} and \texttt{RANLIB=bgranlib}. Thus the easiest thing
to do is to modify all \texttt{Makefile.in} files and the \texttt{configure}
script to set them automatically by \texttt{configure}. One then just
needs to pass the right commands to the \texttt{configure} script:

\noindent \texttt{ ./configure -{}-prefix=/path/to/SPECFEM3DG\_SP
-{}-host=Babel -{}-build=BGP }~\\
 \texttt{ FC=bgxlf90\_r MPIFC=mpixlf90\_r CC=bgxlc\_r AR=bgar ARFLAGS=cru
}~\\
 \texttt{ RANLIB=bgranlib LOCAL\_PATH\_IS\_ALSO\_GLOBAL=false}

@@

\noindent \texttt{bgrun -np 1 -mode VN -exe ./bin/xcreate\_header\_file}

@@

\begin{description}
\item [{\texttt{PROJECT\_NAME}}]
\item [{\texttt{OPTIMIZE\_FOR\_FORTRAN}}] Set to YES
\item [{\texttt{EXTRACT\_ALL}}] Set to YES
\item [{\texttt{EXTRACT\_PRIVATE}}] Set to YES
\item [{\texttt{EXTRACT\_STATIC}}] Set to YES
\item [{\texttt{INPUT}}] From the directory \texttt{specfem3d/doc/Call\_trees}, it is \texttt{"../../src/"}
\item [{\texttt{FILE\_PATTERNS}}] In SPECFEM case, it is \texttt{*.f90* *.F90* *.c* *.cu* *.h*}
\item [{\texttt{HAVE\_DOT}}] Set to YES
\item [{\texttt{CALL\_GRAPH}}] Set to YES
\item [{\texttt{CALLER\_GRAPH}}] Set to YES
\item [{\texttt{DOT\_PATH}}] The path where is located the dot program graphviz (if it is not in your \$PATH)
\item [{\texttt{RECURSIVE}}] This tag can be used to turn specify whether or not subdirectories should be searched for input files as well. In the case of SPECFEM, set to YES.
\item [{\texttt{EXCLUDE}}] Here, you can exclude:  \\
\texttt{"../../src/specfem3D/older\_not\_maintained\_partial\_OpenMP\_port"}\\
\texttt{"../../src/decompose\_mesh/scotch"}\\
\texttt{"../../src/decompose\_mesh/scotch\_5.1.12b"}
\item [{\texttt{DOT\_GRAPH\_MAX\_NODES}}] to set the maximum number of nodes that will be shown in the graph. If the number of nodes in a graph becomes larger than this value, doxygen will truncate the graph, which is visualized by representing a node as a red box. Minimum value: 0, maximum value: 10000, default value: 50.
\item [{\texttt{MAX\_DOT\_GRAPH\_DEPTH}}] to set the maximum depth of the graphs generated by dot. A depth value of 3 means that only nodes reachable from the root by following a path via at most 3 edges will be shown. Using a depth of 0 means no depth restriction. Minimum value: 0, maximum value: 1000, default value: 0.

>>>>

To download the SPECFEM3D\_Cartesian software package, type this:
\begin{verbatim}
git clone --recursive --branch devel https://github.com/geodynamics/specfem3d.git
\end{verbatim}
%
\noindent
We recommend that you add {\texttt{ulimit -S -s unlimited}} to your
{\texttt{.bash\_profile}} file and/or {\texttt{limit stacksize
unlimited }} to your {\texttt{.cshrc}} file to suppress any potential
limit to the size of the Unix stack.\\

Then, to configure the software for your system, run the \texttt{configure}
shell script. This script will attempt to guess the appropriate configuration
values for your system. However, at a minimum, it is recommended that
you explicitly specify the appropriate command names for your Fortran
compiler and MPI package (another option is to define FC, CC and MPIF90
in your .bash\_profile or your .cshrc file):
\begin{verbatim}
./configure FC=ifort MPIFC=mpif90
\end{verbatim}
\noindent
Note that MPI must be installed with MPI-IO enabled because parts of SPECFEM3D perform I/Os through MPI-IO.

Before running the \texttt{configure} script, you should probably
edit file \texttt{flags.guess} to make sure that it contains the best
compiler options for your system. Known issues or things to check
are:
\begin{description}
\item [{\texttt{Intel ifort compiler}}] See if you need to add \texttt{-assume byterecl} for your machine. \textbf{In the case of that compiler, we have noticed that versions dot zero sometimes have bugs or issues that can lead to wrong results when running the code, thus we \emph{strongly} recommend using versions dot one or above (for instance version 13.1 instead of 13.0, version 14.1 instead of 14.0 and so on)}.
\item [{\texttt{IBM compiler}}] See if you need to add \texttt{-qsave}
or \texttt{-qnosave} for your machine.
\item [{\texttt{Mac OS}}] You will probably need to install \texttt{XCODE}.
\end{description}
When compiling on an IBM machine with the \texttt{xlf} and \texttt{xlc}
compilers, we suggest running the \texttt{configure} script with the
following options:

{\footnotesize
\begin{verbatim}
./configure FC=xlf90_r MPIFC=mpif90 CC=xlc_r CFLAGS="-O3 -q64" FCFLAGS="-O3 -q64" -with-scotch-dir=...
\end{verbatim}
}

@@

The SCOTCH library \citep{PeRo96} provides efficent static mapping,
graph and mesh partitioning routines. SCOTCH is a free software package
developed by Fran\c{c}ois Pellegrini et al. from LaBRI and INRIA in Bordeaux,
France, downloadable from the web page \url{https://gforge.inria.fr/projects/scotch/}.
In case no SCOTCH libraries can be found on the system, the configuration
will bundle the version provided with the source code for compilation.
The path to an existing SCOTCH installation can to be set explicitly
with the option \texttt{-{}-with-scotch-dir}. Just as an example:
\begin{verbatim}
./configure FC=ifort MPIFC=mpif90 --with-scotch-dir=/opt/scotch
\end{verbatim}
\noindent
If you use the Intel ifort compiler to compile the code, we recommend
that you use the Intel icc C compiler to compile Scotch, i.e., use:
\begin{verbatim}
./configure CC=icc FC=ifort MPIFC=mpif90
\end{verbatim}

@@

The DO-loop using OpenMP threads has a SCHEDULE property. The \texttt{OMP\_SCHEDULE}
environment variable can set the scheduling policy of that DO-loop.
Tests performed by Marcin Zielinski at SARA (The Netherlands) showed
that often the best scheduling policy is DYNAMIC with the size of
the chunk equal to the number of OpenMP threads, but most preferably
being twice as the number of OpenMP threads (thus chunk size = 8 for
4 OpenMP threads etc). If \texttt{OMP\_SCHEDULE} is not set or is empty, the
DO-loop will assume generic scheduling policy, which will slow down
the job quite a bit.

@@

\noindent Edit file \texttt{flags.guess} and put this for \texttt{FLAGS\_CHECK}:
\begin{verbatim}
-g -qfullpath -O2 -qsave -qstrict -qtune=qp -qarch=qp -qcache=auto -qhalt=w
-qfree=f90 -qsuffix=f=f90 -qlanglvl=95pure -Q -Q+rank,swap_all -Wl,-relax
\end{verbatim}

@@

\noindent Then, to configure the code, type this:
\begin{verbatim}
./configure FC=bgxlf90_r MPIFC=mpixlf90_r CC=bgxlc_r LOCAL_PATH_IS_ALSO_GLOBAL=true
\end{verbatim}

@@

\noindent To compile the code on an IBM BlueGene, Laurent L\'eger from
IDRIS, France, suggests the following: compile the code with
\begin{verbatim}
FLAGS_CHECK="-O3 -qsave -qstrict -qtune=auto -qarch=450d -qcache=auto \
  -qfree=f90 -qsuffix=f=f90 -g -qlanglvl=95pure -qhalt=w -Q \
  -Q+rank,swap_all -Wl,-relax"
\end{verbatim}

@@

\noindent Also, \texttt{AR=ar, ARFLAGS=cru} and \texttt{RANLIB=ranlib}
are hardwired in all \texttt{Makefile.in} files by default, but to
cross-compile on BlueGene/P one needs to change these values to \texttt{AR=bgar,
ARFLAGS=cru} and \texttt{RANLIB=bgranlib}. Thus the easiest thing
to do is to modify all \texttt{Makefile.in} files and the \texttt{configure}
script to set them automatically by \texttt{configure}. One then just
needs to pass the right commands to the \texttt{configure} script:
\begin{verbatim}
./configure --prefix=/path/to/SPECFEM3DG_SP --host=Babel --build=BGP \
  FC=bgxlf90_r MPIFC=mpixlf90_r CC=bgxlc_r AR=bgar ARFLAGS=cru \
  RANLIB=bgranlib LOCAL_PATH_IS_ALSO_GLOBAL=false
\end{verbatim}

@@

\noindent 
\begin{verbatim}
bgrun -np 1 -mode VN -exe ./bin/xcreate_header_file
\end{verbatim}

@@

\begin{description}
\item [{\texttt{PROJECT\_NAME}}]
\item [{\texttt{OPTIMIZE\_FOR\_FORTRAN}}] Set to YES
\item [{\texttt{EXTRACT\_ALL}}] Set to YES
\item [{\texttt{EXTRACT\_PRIVATE}}] Set to YES
\item [{\texttt{EXTRACT\_STATIC}}] Set to YES
\item [{\texttt{INPUT}}] From the directory \texttt{specfem3d/doc/Call\_trees}, it is \texttt{"../../src/"}
\item [{\texttt{FILE\_PATTERNS}}] In SPECFEM case, it is \texttt{*.f90* *.F90* *.c* *.cu* *.h*}
\item [{\texttt{HAVE\_DOT}}] Set to YES
\item [{\texttt{CALL\_GRAPH}}] Set to YES
\item [{\texttt{CALLER\_GRAPH}}] Set to YES
\item [{\texttt{DOT\_PATH}}] The path where is located the dot program graphviz (if it is not in your \$PATH)
\item [{\texttt{RECURSIVE}}] This tag can be used to turn specify whether or not subdirectories should be searched for input files as well. In the case of SPECFEM, set to YES.
\item [{\texttt{EXCLUDE}}] Here, you can exclude:
\begin{verbatim}
../../src/specfem3D/older_not_maintained_partial_OpenMP_port
../../src/decompose_mesh/scotch
../../src/decompose_mesh/scotch_5.1.12b
\end{verbatim}
\item [{\texttt{DOT\_GRAPH\_MAX\_NODES}}] to set the maximum number of nodes that will be shown in the graph. If the number of nodes in a graph becomes larger than this value, doxygen will truncate the graph, which is visualized by representing a node as a red box. Minimum value: 0, maximum value: 10000, default value: 50.
\item [{\texttt{MAX\_DOT\_GRAPH\_DEPTH}}] to set the maximum depth of the graphs generated by dot. A depth value of 3 means that only nodes reachable from the root by following a path via at most 3 edges will be shown. Using a depth of 0 means no depth restriction. Minimum value: 0, maximum value: 1000, default value: 0.
